Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-22 E-Hypergraphs-LICS24.tex
Spacing fixed, proofs aligned

\documentclass[sigconf]{acmart}

@@

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalized as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.

% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & \mathcal K \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal G^{\urcorner}     & \mathcal C \arrow[l] \arrow[r] & ^{\ulcorner} \mathcal H
% \end{tikzcd}
% \]
% CHRIS: WHY IS J DISPLAYING AND NOT R? 
\[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPO_square}
    }
\]

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  We now make this identification throughout without further mention.  Thus we reach the following variant of DPO rewriting. 
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
%                   & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
% \end{tikzcd}
% \]
\[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square_Hyp}
    }
\]
\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acylcicity.

@@

\begin{definition}[E-hypergraphs]
\label{def:e-homo}    
An \emph{e-hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l,<,\consistency)$,  where $(V,E,s,t)$ is an unlabelled hypergraph;  $l :E \to \Sigma + 1$ is a labelling function modified to include an extra value $\bot$ in its codomain; $<$ is a strict partial order on $V + E$ called the \textit{child relation}.  We define the following additional notation.  For $x \in V + E$,  the set of \emph{parents} and \emph{children} of $x$,  respectively: 
	[x) = \{x' ~|~ x' <_c x\; \} \qquad (x] = \{x' ~|~  x <_{c} x'\}
We will denote the unique immediate parent $e$ of $x$ by writing $e <^{\mu} x$.
We call edges $e$ such that $l(e) = \bot$ \textit{hierarchical},  and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \emph{top-level}.
We require that the child relation satisfies the following conditions:
\begin{enumerate}
	\item each parent set contains exclusively hierarchical edges;
	\item Each $x$ has at most one immediate parent;
	\item if $v \in s(e)$ then $e' <^{\mu} e$ iff $e' <^{\mu} v$.
\end{enumerate}
We additionally define an equivalence relation $\consistency_p$,  called the \textit{consistency relation},  on each set $\{x \in V + E ~|~ p <^\mu c\}$ of elements which share the same parent.  
We collect the union of all these equivalence relations into on relation $\consistency = \bigcup_{p} \consistency_p$,  which is defined on a subset of $V+E$. 
The consistency equivalence relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ then $v \consistency e$. 
\label{def:consistency_properties}
Intuitively,  the child relation equips the hypergraph with a hierarchical structure: certain edges,  labelled with $\bot$,  are meant to interpret the dashed e-boxes of  string diagrams.  
These edges can be designated as "parents" to sub-e-hypergraphs,  which interpret the various components of the e-boxes.  These components are distinguished by the consistency relation, which partitions the maximal sub-e-hypergraph below the hierarchical edge.  
Technically,  condition $(3)$ on the child relations says that $<^{\mu}$ respects connectivity of edges and vertices,  and similarly for conditions $(1)$ of the consistency relation. 
\begin{definition}[E-hypergraph homomorphism]
An \textit{e-hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
\begin{enumerate}
    \item $\phi$ is a hypergraph homomorphism
    \item for $v \in V_{\mathcal{F}}$,  if $v$ is not top-level then 
        \phi_{E}(<_{\mathcal{F}}^{\mu}(v)) = <_{\mathcal{G}}^{\mu}(\phi_{V}(v))
        and for $e \in E_{\mathcal{F}}$,  if $e$ is not top-level then 
        \phi_{E}(<_{\mathcal{F}}^{\mu}(e)) = <_{\mathcal{G}}^{\mu}(\phi_{E}(e))  
    % \item
    % $x_1 \#_{\mathcal{F}} x_2$ iff $\phi(x_1) \#_{\mathcal{G}} \phi(x_2)$, where $x_1, x_2 \in \mathcal{F}$
        \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 \consistency_{\mathcal{F}} x_2$ implies $\phi(x_1) \consistency_{\mathcal{G}} \phi(x_2)$. 
%    Or, written functionally, when $x \in E_{\mathcal{F}}$
%    \[
%        \langle \phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}} \rangle^{*}(\consistency_{\mathcal{F}}^{\mu}(i_{E_{\mathcal{F}}}(x)))
%        \subseteq
%        \consistency_{\mathcal{G}}^{\mu}(\phi_{E};i_{E_{\mathcal{G}}}(x))
%    \]
%    , where $\phi_{V};i_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$%, and similarly for $\phi_{E};i_{E_\mathcal{G}}$ so that $\langle \phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}} \rangle : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
%    And, when $x \in V_{\mathcal{F}}$
%    \[
%        \langle \phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}} \rangle^{*}(\consistency_{\mathcal{F}}^{\mu}(i_{V_{\mathcal{F}}}(x)))
%        \subseteq
%        \consistency_{\mathcal{G}}^{\mu}(\phi_{V};i_{V_{\mathcal{G}}}(x))
%    \]
    % Or, on per-component basis, if $i_{E_{\mathcal{F}}}(x_1) \in \hashtag_{\mathcal{F}}(i_{E_{\mathcal{F}}}(x_2))$ then $i_{E_{\mathcal{G}}}(\phi_{E}(x_1)) \in \hashtag_{\mathcal{G}}(i_{E_{\mathcal{G}}}(\phi_{E}(x_2)))$ and similarly for other types of $x_1,x_2$.
\end{enumerate}
If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
The conditions on e-hypergraph homomorphisms require to preserve immediate parents and the conflict relation.  
\begin{definition}[Category of e-hypergraphs]
The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  

@@

\begin{figure}
    \[
    \scalebox{0.55}{
    \tikzfig{combinatorial_semantics/f_plus_g_new}
    }
    \]
    \caption{$+$ of two morphisms in $\Ecospans$}
    \label{fig:A+B}
\end{figure}
Recall from Section \ref{sec:e-hypergraphs} that $ \WellTypedMdaEcospans$ has a symmetric monoidal structure inherited from the coproduct structure of $\textbf{EHypI}(\Sigma)$.  
Recall further that typed $\Sigma^+$-terms are those categorical combinator terms freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sigma_{1,1}$, $(;\!)$ and $\otimes$,  and $f+g$.  We will write $\Sigma^+(n,m)$ for the set of $\Sigma^+$-terms of type $n \to m$.

@@

            \item When $x$ is not a top-level vertex, 
                \[
                \phi_{E}(<_{\mathcal{F}}^{\mu}(\iota_{V_{\mathcal{F}}}(x))) = <_{\mathcal{G}}^{\mu}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
                \]
                and
                \[
                \phi_{E}(<_{\mathcal{F}}^{\mu}(\iota_{E_{\mathcal{F}}}(x))) = <_{\mathcal{G}}^{\mu}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))  
                \] when $x$ is a not top-level edge.          
                \item
            When $x \in E_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}} ]^{*}(\consistency_{\mathcal{F}}(\iota_{E_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
            \]
            where $\phi_{V};\iota_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$, and similarly for $\phi_{E};\iota_{E_\mathcal{G}}$ so that $[\phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}}] : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
            And, when $x \in V_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{F}}(\iota_{V_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
            \]
            \end{enumerate}
    \end{definition}
\end{remark}

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{discrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{c,X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
        \\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
    % \begin{align*}
    % S_{E} &= \{\\
    % &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    % &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    % &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    % &\}
    % \end{align*}
and let relations $R_{V},R_{E}$ be their reflexive symmetric and transitive closures respectively.

@@

    \item If there exists $v' \in V_{X+Y}$ such that $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v')) \not = \varnothing$ and $v' \sim v$
    \[
        \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]_{V})) = [[]_{V}^{\consistency}, []_{E}^{\consistency}]^{*}(\bigcup_{v' \in V_{X} + V_{Y} | [v']_{V} = [v]_{V}} \consistency_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v')))
    \]
    \item If for all $v' \in V_{X+Y}$ such that $v' \sim v$, $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v')) = \varnothing$ and there exists $[u]$ such that there is an undirected path from $[v]$ to $[u]$ and $\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}([u])) \not = \varnothing$, then
    \[
        \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]_{V})) = \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{E_{X} + E_{Y}}([u]_{V}))
    \]
    \end{enumerate}
    where $\iota_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    % We also build this relation step by step as the relation above: first steps (1) for nodes and edges and then we interleave steps for edges and for nodes.

>>>>

\documentclass[sigconf,9pt]{acmart}

@@

The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalized as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Figure~\ref{fig:dpo_dpoi}~(left).

% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & \mathcal K \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal G^{\urcorner}     & \mathcal C \arrow[l] \arrow[r] & ^{\ulcorner} \mathcal H
% \end{tikzcd}
% \]
% CHRIS: WHY IS J DISPLAYING AND NOT R? 
% \[
%     \scalebox{0.8}{
%         \tikzfig{combinatorial_semantics/DPO_square}
%     }
% \]

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Figure~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
% \[
% \begin{tikzcd}
% \mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
% \mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
%                   & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
% \end{tikzcd}
% \]
\begin{figure}[t!]
    \begin{subfigure}{0.45\linewidth}
    \[
            \scalebox{0.8}{
                \tikzfig{combinatorial_semantics/DPO_square}
            }
    \] 
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\linewidth}
        \[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square_Hyp}
    }
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acylcicity.

@@

\begin{definition}[E-hypergraphs]
\label{def:e-homo}    
An \emph{e-hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l,<,\consistency)$,  where $(V,E,s,t)$ is an unlabelled hypergraph;  $l :E \to \Sigma + 1$ is a labelling function modified to include an extra value $\bot$ in its codomain; $<$ is a strict partial order on $V + E$ called the \textit{child relation}.  We define the following additional notation.  For $x \in V + E$,  the set of \emph{parents} and \emph{children} of $x$,  respectively: 
\[
	[x) = \{x' ~|~ x' <_c x\; \} \qquad (x] = \{x' ~|~  x <_{c} x'\}
\]
We will denote the unique immediate parent $e$ of $x$ by writing $e <^{\mu} x$.
We call edges $e$ such that $l(e) = \bot$ \textit{hierarchical},  and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \emph{top-level}.
We require that the child relation satisfies the following conditions:
\begin{enumerate}
	\item each parent set contains exclusively hierarchical edges;
	\item Each $x$ has at most one immediate parent;
	\item if $v \in s(e)$ then $e' <^{\mu} e$ iff $e' <^{\mu} v$.
\end{enumerate}
We additionally define an equivalence relation $\consistency_p$,  called the \textit{consistency relation},  on each set $\{x \in V + E ~|~ p <^\mu c\}$ of elements which share the same parent.  
We collect the union of all these equivalence relations into on relation $\consistency = \bigcup_{p} \consistency_p$,  which is defined on a subset of $V+E$. 
The consistency equivalence relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ then $v \consistency e$. 
\label{def:consistency_properties}
\end{definition}
Intuitively,  the child relation equips the hypergraph with a hierarchical structure: certain edges,  labelled with $\bot$,  are meant to interpret the dashed e-boxes of  string diagrams.  
These edges can be designated as "parents" to sub-e-hypergraphs,  which interpret the various components of the e-boxes.  These components are distinguished by the consistency relation, which partitions the maximal sub-e-hypergraph below the hierarchical edge.  
Technically,  condition $(3)$ on the child relations says that $<^{\mu}$ respects connectivity of edges and vertices,  and similarly for conditions $(1)$ of the consistency relation. 
\begin{definition}[E-hypergraph homomorphism]
An \textit{e-hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
\begin{enumerate}
    \item $\phi$ is a hypergraph homomorphism
    \item for $v \in V_{\mathcal{F}}$,  if $v$ is not top-level then 
	\[
        \phi_{E}(<_{\mathcal{F}}^{\mu}(v)) = <_{\mathcal{G}}^{\mu}(\phi_{V}(v))
        \] 
        and for $e \in E_{\mathcal{F}}$,  if $e$ is not top-level then 
        \[
        \phi_{E}(<_{\mathcal{F}}^{\mu}(e)) = <_{\mathcal{G}}^{\mu}(\phi_{E}(e))  
        \] 
        \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 \consistency_{\mathcal{F}} x_2$ implies $\phi(x_1) \consistency_{\mathcal{G}} \phi(x_2)$. 
\end{enumerate}
If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
\end{definition}
The conditions on e-hypergraph homomorphisms require to preserve immediate parents and the conflict relation.  
\begin{definition}[Category of e-hypergraphs]
The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
\end{definition}

@@

\begin{figure}
    \[
    \scalebox{0.55}{
    \tikzfig{combinatorial_semantics/f_plus_g_new}
    }
    \]
    \captionsetup{belowskip=-4ex}
    \caption{$+$ of two morphisms in $\Ecospans$}
    \label{fig:A+B}
\end{figure}
Recall from Section \ref{sec:e-hypergraphs} that $ \WellTypedMdaEcospans$ has a symmetric monoidal structure inherited from the coproduct structure of $\textbf{EHypI}(\Sigma)$.  
Recall further that typed $\Sigma^+$-terms are those categorical combinator terms freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sigma_{1,1}$, $(;\!)$ and $\otimes$,  and $f+g$.  We will write $\Sigma^+(n,m)$ for the set of $\Sigma^+$-terms of type $n \to m$.

@@

            \item When $x$ is not a top-level vertex, 
                \[
                \phi_{E}(<_{\mathcal{F}}^{\mu}(\iota_{V_{\mathcal{F}}}(x))) = <_{\mathcal{G}}^{\mu}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
                \]
                and
                \[
                \phi_{E}(<_{\mathcal{F}}^{\mu}(\iota_{E_{\mathcal{F}}}(x))) = <_{\mathcal{G}}^{\mu}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))  
                \] when $x$ is a not top-level edge.          
                \item
            When $x \in E_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}} ]^{*}(\consistency_{\mathcal{F}}(\iota_{E_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
            \]
            where $\phi_{V};\iota_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$, and similarly for $\phi_{E};\iota_{E_\mathcal{G}}$ so that $[\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}] : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
            And, when $x \in V_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{F}}(\iota_{V_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
            \]
            \end{enumerate}
    \end{definition}
\end{remark}

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{discrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
        \\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
    % \begin{align*}
    % S_{E} &= \{\\
    % &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    % &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    % &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    % &\}
    % \end{align*}
and let relations $R_{V},R_{E}$ be their reflexive symmetric and transitive closures respectively.

@@

    \item If there exists $v' \in V_{X+Y}$ such that $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v')) \not = \varnothing$ and $v' \sim v$
    \[
        \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]_{V})) = [[]_{V}^{\consistency}, []_{E}^{\consistency}]^{*}(\bigcup_{v' \in V_{X} + V_{Y} | [v']_{V} = [v]_{V}} \consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v')))
    \]
    \item If for all $v' \in V_{X+Y}$ such that $v' \sim v$, $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v')) = \varnothing$ and there exists $[u]$ such that there is an undirected path from $[v]$ to $[u]$ and $\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}([u])) \not = \varnothing$, then
    \[
        \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]_{V})) = \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{E_{X} + E_{Y}}([u]_{V}))
    \]
    \end{enumerate}
    where $\iota_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    % We also build this relation step by step as the relation above: first steps (1) for nodes and edges and then we interleave steps for edges and for nodes.
